Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-02 appendix.tex
merge with youdin's updates
additional changes to sec 5.3, added new appendix D2, minor addition to appendix C2.2

\section{Derivation of the approximate equations}\label{adia_improve}
Here we detail the derivation of Eq. \ref{nearly_iso_explicit} used in the 
analytical derivations of \S\ref{analytical}.  Starting with Eqs.\ \ref{lin_mass} --- \ref{lin_energy},
we set $\Gamma = 1$ and $\zeta = 0$ for the vertically isothermal limit and the fully-radially-local 
approximation, respectively.  We  eliminate
the horizontal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{subequations}\label{A1}\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{\p\ln{\rho}}{\p z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{\p\Omega^2}{\p z} -
    \frac{\p \ln{\rho}}{\p z}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{\p\ln{\rho}}{\p z}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}\end{subequations}
where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2.
\end{align} 
%

@@

We  eliminate $W,\, Q$ from Eqs.\ \ref{A1}, using  Eqs. \ref{chi} and \ref{dkappa2} to obtain  
\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z, 
\end{align}
where we have replaced $\p/\p z$ by $d/dz$ for the
fully-radially-local treatment. Now we make the low frequency and Keplerian
approximations, setting $D\to 
\OmK^2$, to give  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3},
\end{align} 
in terms of dimensionless variables of Eq.\ \ref{nondim}.

Retaining terms to first order in the disk aspect ratio, $h \ll 1$, gives
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z.\label{vertiso_gov_nondim}
\end{align}
Approximating the density field by Eq.\ \ref{rhoisothin} then gives 
Eq. \ref{nearly_iso_explicit}.

@@

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\upsilon}^2\khat^2 =
  \left(L-\hat{\upsilon}^2\right)\left(1-\hat{\upsilon}^2\right), \label{iso_disp_full}
\end{align}
which is just the dispersion relation for axisymmetric isothermal waves in a
vertically isothermal disk (e.g. \citealt{okazaki87}; \citealt{takeuchi98}; \citealt{tanaka02}; 
\citealt{zhang06}; \citealt{ogilvie13}; \citealt{barker14}; \citetalias{barker15}). In 
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\upsilon} = \hat{\omega}$ is real and the disk
is stable. The low frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$.  This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.

@@

\subsection{Finding marginal stability}\label{disp_neut_limit}
To investigate marginal stability we set $\hat{\sigma} = 0$ so the frequency, $\hat{\upsilon}=\hat{\omega}_c$, is real.  
We take the short radial wavelength limit, $\khat^2\gg 1$, of the coefficients in Eq.\ \ref{coeffs}.  The real and 
imaginary parts of Eq. \ref{relax_disp} then give relations for $\hat{\omega}_c$ and $\beta_c$, the cooling time for 
marginal stability, as 
\begin{align}\
  0 =&M(M+1)(1-h^2 q^2 \khat^2) + 4\beta_c h q M(M+1)\left(\hat{\omega}_c\khat\right) 
 +\left[1 +
    \gamma\beta_c^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}_c\khat\right)^2 \label{relax_cond1} \\
 &+ 2h q \gamma\beta_c \left(\hat{\omega}_c\khat\right)^3 -  \left(\hat{\omega}_c\khat\right)^4 
  + \beta_c^2\gamma^2\hat{\omega}_c^2
  \left(\hat{\omega}_c\khat\right)^4, \notag \\
   0=& 2h q M (M+1)\khat +
   \beta_c(1-\gamma)(1+2M)^2\khat\left(\hat{\omega}_c\khat\right) 
   + h q \khat \left(\hat{\omega}_c\khat\right)^2 -
   2\beta_c\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^2
   - h q
   \gamma^2\beta_c^2\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^3 \label{relax_cond2} \\
   &+
   2\beta_c\gamma\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^4 \notag. 
\end{align}
We next (re)-apply the low frequency and thin approximations, $h, \hat{\omega}_c \ll 1$ and 
further apply the rapid cooling limit $\beta_c \ll 1$.  We retain terms with the combinations 
$\hat{\omega}_c \hat{k}$, $\beta_c/h$, and $h \hat{k}$ which could be order unity.  For inertial waves we know that
$\hat{\omega}_c \sim 1/\hat{k}$ holds, while $\beta_c/h \sim O(1)$ will be justified \emph{a posteriori}
and the magnitude of $h \hat{k}$ depends on the choice of wavenumber.  Applying this ordering we 
arrive at Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}.

>>>>

\section{Derivation of the approximate equations}\label{adia_improve}
Here we detail the derivation of Eq. \ref{nearly_iso_explicit} used in the 
analytical discussion of \S\ref{analytical}.  Starting with Eqs.\ \ref{lin_mass} --- \ref{lin_energy},
we set $\Gamma = 1$ and $\zeta = 0$ for the vertically isothermal limit and the fully-radially-local 
approximation, respectively.  We  eliminate
the horizontal velocity perturbations ($\delta v_x,\, \delta v_y$) to
obtain  
%we can eliminate variables 
%to obtain a pair or ordinary differential
%equations for $W,Q$ as 
\begin{subequations}\label{A1}\begin{align}
  & \ii\upsilon\delta v_z = \frac{dW}{dz} + \frac{\p\ln{\rho}}{\p z}\left(W-Q\right),\label{ode_vz} \\
  &\frac{\upsilon^2}{c_s^2}Q + \frac{\upsilon^2k_x^2}{D}W =\ii\upsilon \left(\frac{\ii
      k_x r}{D}\frac{\p\Omega^2}{\p z} -
    \frac{\p \ln{\rho}}{\p z}\right)\delta v_z - \ii\upsilon \frac{d\delta v_z}{dz},
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right] 
  % \notag\\
  % &- \frac{d^2W}{dz^2} - \frac{d\ln{\rho}}{dz}\left(\frac{dW}{dz} -
  %   \frac{dQ}{dz}\right) - \frac{d^2\ln{\rho}}{dz^2}\left(W-Q\right),
  \label{ode_w}\\
%  &\sigma^2W - \frac{\gamma}{\Gamma}\sigma^2 Q +
%  \frac{\ii\sigma}{t_c}\left(W-\frac{Q}{\Gamma}\right) 
%  =\ii \sigma c_s^2\frac{d\ln{\rho}}{dz}\left(\frac{\gamma}{\Gamma} -
%    1\right)\delta v_z,  
  &\upsilon^2W - \gamma\upsilon^2 Q +
  \frac{\ii\upsilon}{t_c}\left(W-Q\right) 
  =\ii \upsilon c_s^2\frac{\p\ln{\rho}}{\p z}\left(\gamma -
    1\right)\delta v_z,
  % 
  % \left[\frac{dW}{dz} + \frac{d\ln{\rho}}{dz}\left(W-Q\right)\right],
  \label{ode_Q} 
\end{align}\end{subequations}
where
\begin{align}
  D \equiv \kappa^2 - \upsilon^2.
\end{align} 
%

@@

We  eliminate $W,\, Q$ from Eqs.\ \ref{A1}, using  Eqs. \ref{chi} and \ref{dkappa2} to obtain  
\begin{align}\label{full_ode}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left[1 + \frac{\ii k_x c_s^2
      q}{Dr} - \frac{k_x^2c_s^2}{\left(k_x^2c_s^2 + \chi
        D\right)}\frac{qc_s^2F}{Dr^2}\right]\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left\{\upsilon^2\left(\frac{k_x^2}{D} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{Dr}\right)\frac{d^2\ln\rho}{dz^2} -
    \frac{c_s^2}{D}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left[\left(1-\chi\right) +
     \frac{\chi}{\left(k_x^2c_s^2 + \chi D\right)}\frac{qc_s^2 F}{r^2}\right] 
   \right\}\delta v_z, 
\end{align}
where we have replaced $\p/\p z$ by $d/dz$ for the
fully-radially-local treatment. Now we make the low frequency and Keplerian
approximations, setting $D\to 
\OmK^2$, to give  
\begin{align}
  &\delta v_z ^{\prime\prime} + \left[1 + \ii h q\hat{k} -
    \frac{ \hat{k}^2}{
      \left(\hat{k}^2+\chi\right)}q h^2F\right]\ln\rho^{\prime}\delta v_z^\prime +
  \left\{\left(\chi + \ii h q
      \hat{k}\right)\ln\rho^{\prime\prime} - \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii h
      q\hat{k}\right)\left[1 - \chi +
      \frac{\chi}{\left(\hat{k}^2+\chi\right)}qh^2F\right]\right\}\delta v_z \notag\\&=
  -\hat{\upsilon}^2\left(\hat{k}^2+\chi\right)\delta v_z\label{adia_diso3},
\end{align} 
in terms of dimensionless variables of Eq.\ \ref{nondim}. 
Retaining terms to first order in the disk aspect ratio, $h \ll 1$, gives
\begin{align}
  0= \delta v_z ^{\prime\prime} + \left(1 + \ii h
     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime 
   +
   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}
     - \ln\rho^{\prime
       2}\left(\khat^2 -
       \ii h
       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z.\label{vertiso_gov_nondim}
\end{align}
Approximating the density field by Eq.\ \ref{rhoisothin} then gives 
Eq. \ref{nearly_iso_explicit}.

@@

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\upsilon}^2\khat^2 =
  \left(L-\hat{\upsilon}^2\right)\left(1-\hat{\upsilon}^2\right), \label{iso_disp_full}
\end{align}
which is just the dispersion relation for axisymmetric isothermal waves in a
vertically isothermal disk (e.g. \citealt{okazaki87}; \citealt{takeuchi98}; \citealt{tanaka02}; 
\citealt{zhang06}; \citealt{ogilvie13}; \citealt{barker14}; \citetalias{barker15}). In 
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\upsilon} = \hat{\omega}$ is real and the disk
is stable. The low frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$. This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}. 
Note that $L=M+1$, where $M\geq 0$ is the mode number used for analytical
discussion (\S\ref{analytical}) based on the reduced equation for
$\delta v_z$, rather than for $W$ as considered here. 
%Note that for
%analytical discussions (\S\ref{analytical}) we work with an equation
%for $\delta v_z$ so that $L=M+1$.

@@

\subsection{Finding marginal stability}\label{disp_neut_limit}
To investigate marginal stability we set $\hat{\sigma} = 0$ so the frequency, $\hat{\upsilon}=\hat{\omega}_c$, is real.  
We take the short radial wavelength limit, $\khat^2\gg 1$, of the coefficients in Eq.\ \ref{coeffs}.  The real and 
imaginary parts of Eq. \ref{relax_disp} then give relations for $\hat{\omega}_c$ and $\beta_c$, the cooling time for 
marginal stability, as 
\begin{align}\
  0 =&M(M+1)(1-h^2 q^2 \khat^2) + 4\beta_c h q M(M+1)\left(\hat{\omega}_c\khat\right) 
 +\left[1 +
    \gamma\beta_c^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}_c\khat\right)^2 \label{relax_cond1} \\
 &+ 2h q \gamma\beta_c \left(\hat{\omega}_c\khat\right)^3 -  \left(\hat{\omega}_c\khat\right)^4 
  + \beta_c^2\gamma^2\hat{\omega}_c^2
  \left(\hat{\omega}_c\khat\right)^4, \notag \\
   0=& 2h q M (M+1)\khat +
   \beta_c(1-\gamma)(1+2M)^2\khat\left(\hat{\omega}_c\khat\right) 
   + h q \khat \left(\hat{\omega}_c\khat\right)^2 -
   2\beta_c\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^2
   - h q
   \gamma^2\beta_c^2\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^3 \label{relax_cond2} \\
   &+
   2\beta_c\gamma\hat{\omega}_c\left(\hat{\omega}_c\khat\right)^4 \notag. 
\end{align}
We recall that in the low frequency and thin-disk approximations, $h,
|\hat{\omega}_c| \ll 1$. We note that $|q|=O(1)$ and 
$\gamma>1$ but is $O(1)$. We consider $M$ to be $O(1)$ or smaller, so 
that $|\hat{\omega_c}\khat|=O(1)$, since for inertial waves
$|\hat{\omega}\khat|^2\simeq 1 + M $. Finally, we assume 
$\beta_c \ll 1$, to be justified \emph{a posteriori}. Applying these
approximations give
Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}.

\subsection{Marginal stability for the fundamental mode}\label{bcrit_alt}
If we apply the above approximations for the fundamental mode $M=0$ to
Eq. \ref{relax_cond1}---\ref{relax_cond2}, but with the less
restrictive assumption $\beta_c \lesssim O(1)$, we expect the balance  
\begin{subequations}\begin{align}
    &0  = \left[1 + \gamma\beta_c^2\left(1-\gamma\right)\right] + 2h q
    \gamma \beta_c \left(\hat{\omega}_c\khat\right) - \left(\hat{\omega}_c\khat\right)^2,\\
    &0 =  \beta_c(1-\gamma) + hq \left(\hat{\omega}_c\khat\right). 
  \end{align}\end{subequations}
Then
\begin{align}
  \beta_c = \frac{h|q|}{\gamma-1}\left[1 - \frac{\gamma h^2 
      q^2}{\gamma-1}\right]^{-1/2},
\end{align}
which reduces to Eq. \ref{iso_vsi_cond} for thin-disks ($h\ll1$). We
thus deduce that in fact $\beta_c\ll 1$, which motivates us to make this
strong assumption from the onset.

### 2015-05-02 linear_adia.tex
very minor tweaks to 5 (done for now) and starting with 6.
update with youdin's update
additional changes to sec 5.3, added new appendix D2, minor addition to appendix C2.2

\subsection{An upper limit to the cooling time}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta_c$ that allows growth the VSI.  In Appendix \ref{disp_neut_limit}
we derive the following relation between $\beta_c$ and the wave frequency,
$\hat{\omega}_c$  at marginal stability to the VSI:
\begin{subequations}\begin{align}
 \left(\hat{\omega}_c\khat\right)^4  = &  
% \left[1 + \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}_c\khat\right)^2 \label{relax_cond_simp1}\\ 
 \left(\hat{\omega}_c\khat\right)^2  + M(M+1)\left(1-h^2q^2\khat^2\right) , \label{relax_cond_simp1} \\
   \left(\hat{\omega}_c\khat\right)^2 = & \frac{\beta}{h q} (\gamma-1)(1+2M)^2
   \left(\hat{\omega}_c\khat\right) \label{relax_cond_simp2}\\
   &- 2M(M+1) . \notag
\end{align}\end{subequations}
These equations hold for $\beta_c \ll 1$ and $\khat \sim 1/\hat{\omega}_c \gg1$,
as explained in Appendix \ref{disp_neut_limit}.

Eq. \ref{relax_cond_simp1} gives an approximate result for the frequency
\begin{align}
 \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left[1  + O( M|h q \khat|^2)\right]  \label{marg_freq_approx}
%  \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left(1 -
%    \frac{M|h q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
which holds for either (a) $M= 0$ and arbitrary $|h q \hat{k}|$ or (b) $|h q \hat{k}| \ll 1$ and arbirtary $M \sim O(1)$. 
These distinct conditions mean that our results are most robust for the fundamental vertical mode ($M = 0$) but also
hold for higher order vertical modes when radial wavelengths are not too short.
Note that Eq. \ref{marg_freq_approx} 
reduces (\emph{nearly}?)\footnote{\emph The correspondence is only exact 
if $L = M+1$.  Is this the case (in which case at least note in the appendix) or is there a discrepancy (in which case 
we can't say it reduces)?}
 to the dispersion relation for low-frequency inertial waves 
%in vertically isothermal disks without vertical shear 
(see Appendix\ref{stable_novshear} for details).

Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta_c &\simeq  \frac{h|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}
%  \left[1 + \frac{M(2M-1)|h q   \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
where the above-mentioned approximations still hold, but the correction term ($O(M |h q \khat|^2)$) is now suppressed.
Eq. \ref{bcrit_gen} says that higher $M$ modes have lower $\beta_c$, i.e.\ require faster cooling for VSI growth.
This trend is confirmed numerically in \S\ref{therm_relax_eff} (and the trend holds even when the restrictive assumption
on $|h q \hat{k}|$ is dropped).

We are most interested in the longest cooling time which allows growth for any $M$.  
This criterion is clearly set by the fundamental, $M = 0$, mode.  Our cooling criterion for 
VSI growth is thus 
 \begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{h|q|}{\gamma-1} . 
\end{align}
The thin disk approximation, $h \ll 1$, justifies the $\beta_c \ll 1$ approximation.
 We heuristically explain $\beta_\mathrm{crit}$ in
\S\ref{vsi_require}, and numerically test its validity in \S\ref{bcrit_num_test}.

%Eq. \ref{iso_vsi_cond} suggests that as the thermal timescale $\beta$
%is increased from zero, higher order modes with $M>0$
%will be stabilized before the fundamental mode. We indeed observe this
%trend in numerical calculations (\S\ref{therm_relax_eff}), despite
%adopting somewhat restrictive assumptions for $M>0$ in order to arrive
%at the simple expression Eq. \ref{bcrit_gen}.

%We can interpret  $\beta_\mathrm{crit}$ as the thermal timescale below
%which buoyancy becomes ineffective and perturbations are effectively
%isothermal so the VSI can operate.  We see that in a thin disk, a very short thermal timescale
%is required ($\beta_\mathrm{crit}\ll 1$) because
%$\beta_\mathrm{crit}\propto h$. Physically, this is because
%vertical shear is weak in a thin disk, so even a small entropy
%gradient can significantly reduce VSI growth rates.

>>>>

\subsection{An upper limit to the cooling time}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta_c$ that allows growth of the VSI. In Appendix \ref{disp_neut_limit}
we derive the following relation between $\beta_c$ and the wave
frequency $\hat{\omega}_c$, \emph{assuming} marginal stability to the
VSI:  
\begin{subequations}\begin{align}
    \left(\hat{\omega}_c\khat\right)^4  =& \left(\hat{\omega}_c\khat\right)^2 
  + M(M+1)\left(1-h^2q^2\khat^2\right) ,\label{relax_cond_simp1}  \\
  \left(\hat{\omega}_c\khat\right)^2 = & \frac{\beta_c}{h q} (\gamma-1)(1+2M)^2
  \left(\hat{\omega}_c\khat\right) \label{relax_cond_simp2}\\
  &- 2M(M+1) . \notag
\end{align}\end{subequations}
These equations consider $\khat^2\gg 1$, and assumes 
$|\hat{\omega}\khat| = O(1)$, $M\lesssim O(1)$ and $\beta_c\ll 1$.

Eq. \ref{relax_cond_simp1} admits the following   
approximate solution for the frequency if $M|hq\khat|^2/(2M+1)\lesssim
1$, 
\begin{align}
  % \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left[1  + O( M|h q \khat|^2)\right]  \label{marg_freq_approx}
  \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|h q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}  
% which holds for either (a) $M= 0$ and arbitrary $|h q \hat{k}|$ or (b) $|h q \hat{k}| \ll 1$ and arbirtary $M \sim O(1)$. 
% These distinct conditions mean that our results are most robust for the fundamental vertical mode ($M = 0$) but also
% hold for higher order vertical modes when radial wavelengths are not
% too short. 
Note that Eq. \ref{marg_freq_approx} reduces to the dispersion
relation for low-frequency inertial waves in the absence of vertical
shear (see Appendix \ref{stable_novshear} for details).

If we further assume $|hq\khat|^2\ll 1$, we obtain from
Eq. \ref{relax_cond_simp2}---\ref{marg_freq_approx} that  
\begin{align}\label{bcrit_gen} 
  \beta_c &\simeq  \frac{h|q|}{(\gamma - 1)}\frac{\sqrt{1+M}}{(1+2M)}.   
%  \frac{\sqrt{1+M}}{(1+2M)}
%  \left[1 + \frac{M(2M-1)|h q   \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}%\notag
%>>>>>>> 8816838e4a06974fc61d74b84a2474adaeee3ebc
\end{align}
% but the correction term ($O(M |h q \khat|^2)$) is now suppressed.    
Eq. \ref{bcrit_gen} indicates that higher $M$ modes have lower $\beta_c$,
i.e.\ require faster cooling for VSI growth\footnote{In fact,   
  this conclusion does not require $|hq\khat|^2$ to be very 
  small. The may be seen from the more complete expression for
  $\beta_c$ obtained by substituting Eq. \ref{marg_freq_approx} into  
  Eq. \ref{relax_cond_simp2} and retaining the term proportional to 
  $|hq\khat|^2$.}.    
This trend is confirmed numerically in \S\ref{therm_relax_eff}.

%(and the trend holds even when the restrictive assumption
%on $|h q \hat{k}|$ is dropped).

We are most interested in the longest cooling time which allows growth for any $M$.  
This criterion is clearly set by the fundamental, $M = 0$, mode.  Our
cooling criterion for  
VSI growth is thus  
 \begin{align}\label{iso_vsi_cond}
  \beta_c\leq \beta_\mathrm{crit}  \equiv
  \frac{h|q|}{\gamma-1} . 
\end{align}
The thin disk approximation, $h \ll 1$, indicates that $\beta_c \ll
1$, as assumed to obtain Eq. \ref{marg_freq_approx}. We heuristically
explain $\beta_\mathrm{crit}$ in \S\ref{vsi_require}, and numerically
test its validity in \S\ref{bcrit_num_test}.

In Appendix \ref{bcrit_alt}, we give an alternative derivation for 
$\beta_\mathrm{crit}$ which does not assume it is much less
than unity from the onset. This derivation also independent of $|h q
\khat|$.

% For $M=0$, it is straight foward to solve Eq. \ref{relax_cond_simp1} 
% --- \ref{relax_cond_simp2} exactly and obtain the maximum cooling time
% for the fundamental mode, $\beta_\mathrm{crit}$
% (Eq. \ref{iso_vsi_cond}), which shows that $\beta_\mathrm{crit}\ll 1$
% because we consider thin disks.
